# A (brief) introduction to R  {#Intro_to_R}

How to use R.

## Why code?

Reproducible, repeatable, modifiable

## Here's how to use R 

```{r}
1 + 2
```


## Assigning variables
```{r}
# Assigns values to variables
bob <- 1
bill <- 2

# Addition
bob + bill
# Multiplication
bob * bill
# Division
bob / bill
# Subtraction
bob - bill
```

## Data types

```{r}

# Character
a <- "Bob"

# Numeric
b <- 3.567

# Integer
c <- 3

# Logical
d <- TRUE

```

## Vectors

```{r}

# Creating a list of numbers
ls <- c(1, 2, 3, 4, 5)
ls

# Creating a sequence
sq <- seq(1,5,1)
sq

# Creating a repetitive list
rp <- rep(10, 5)
rp

# Multiplying together
ls * rp

```

## Loading data

```{r}

# Load data
df <- read.csv("C:/Users/44797/Desktop/Repo/GEOG-70581/data/flow_data.csv")
head(df)

# What columns do we have
colnames(df)

# What data do they store?
str(df)

# We can access columns by using the $ operator

# df$Date
# df$Flow


```

This is a dataframe, containing two columns, in a similar format to something you may be familiar with in Excel. It has a column of dates and a column of river flow (measured in m^3 per second).

We may want to index using `df[row.index, column.index]`

```{r}

# We can extract certain rows

# Extract rows 1 to 4, all columns
df[1:4,] 

# Extract rows 1 to 4, just the second column
df[1:4, 2]


```

